Compositional symbolic execution is a technique allowing for the scalable verification of code, using separation logic. Gillian is a compositional symbolic execution engine that has the unique property of being parametric on the state model, allowing verification of any language for which a state model can be defined.

Abstract separation logic is a branch of separation logic that allows reasoning about an abstract state. A state-of-the-art theory in this field is Iris, which introduces the notion of resource algebras, enabling the construction of sound complex state from simple elements.

Previous work introduces state model constructions in Gillian, though these transformers were only described on paper. Furthermore, they were formalised using partial commutative monoids, limiting their applicability and causing soundness issues. Following from such works, we show how resource algebras can be adapted into an abstract verified CSE engine, greatly facilitating the construction of sound state models. We then define a wide range of different state model transformers that can be used to assemble state models that carry the desired soundness properties by construction. We also prove at the theory level that some optimisations of these transformers are sound.

Following this, we implement the defined state model transformers, creating a library of easy to use modules and functors that can be combined to model real world languages in Gillian, while being verifiably sound. We instantiate Gillian to JavaScript, C, and WISL using them, greatly reducing developer effort.

Finally, we evaluate the instantiations built with our library and compare them to previous instantiations of Gillian, highlighting a clear performance improvement for JavaScript and C. This allows us to conclude that our library enables a developer to instantiate Gillian to real world languages more easily, with less code, and with exceptional performance -- all the while guaranteeing soundness.
